# The spectrogram

```{r}
#| echo: false

library(praatpicture)
```


The spectrograms produced by `praatpicture` are by default similar to those produced by Praat, but users have a lot of options regarding both the general appearance of spectrograms and the underlying signal processing. We'll cover those options here.

## Appearance

### Frequency range

The spectrogram by default shows a frequency range between 0--5,000 Hz. The frequency range can be controlled with the `spec_freqRange` argument. For example, here we've removed the lowest frequencies (< 200 Hz), and extended the upper range to 10,000 Hz. This can be useful when you want to see what's going on with sibilant-like sounds!

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_freqRange = c(200,10000))
```

### Dynamic range {#sec-spec-dynrange}

The dynamic range of the spectrogram is set to 50 dB by default, meaning that energy below 50 dB under the maximum amplitude in the spectrogram is rendered as white. This can be controlled with the `spec_dynamicRange` argument. The current Praat default is actually a bit higher, namely 70 dB, as shown here:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_dynamicRange = 70)
```

A very high dynamic range will give a spectrogram without any white areas:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_dynamicRange = 150)
```

And with a very low dynamic range, only areas with very high energy is shown:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_dynamicRange = 30)
```

### Coloring options

By default, areas of the specrogram with low energy are rendered in white and areas with energy in black. This is controlled with the `spec_color` argument, which should be at least two colors. For example, you could render high frequencies in blue like so:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_color = c('white', 'blue'))
```

It is also possible to use more complicated color schemes, by specifying both the color used for lowest values, highest values, and any number of in-between values. For example, this plot uses brown for high values, white for low values, and a bunch of pink-ish hues for in-between values

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_color = c('white', 'pink', 'magenta', 'purple', 'brown'))
```

### Axis label

The label along the y-axis says `Frequency (Hz)` by default. This is controlled with the `spec_axisLabel` argument; here's one that just says `Spectrogram`:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_axisLabel = 'Spectrogram')
```

## Signal processing

### Source signal

When plotting multi-channel data, the spectrogram is always based on one of the channels. By default it is the first one, as shown here:

```{r}
#| fig-align: center

praatpicture('ex/multichannel.wav',
             start = 0.6, end = 2.4,
             frames = c('sound', 'spectrogram'))
```

If we were interested in using a different channel for the spectrogram, the `spec_channel` argument can be used for that purpose. Here's a spectrogram showing the intraoral pressure signal:

```{r}
#| fig-align: center

praatpicture('ex/multichannel.wav',
             start = 0.6, end = 2.4,
             frames = c('sound', 'spectrogram'),
             spec_channel = 4)
```

### Window length

The default window length used for generating spectra is 5 ms, following Praat. This gives you a narrowband spectrogram. Window length is controlled with the `spec_windowLength` argument. For example, using 30 ms will give you a broadband spectrogram. This has lower temporal resolution, but is great for seeing the harmonics!

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowLength = 0.03)
```

### Window shape

Spectra are generated with Gaussian windows by default, but a range of other shapes are available. A bunch of other options are available, although the difference between most of them is marginal. Gaussian window shapes are a bit slower, but the difference is small enough that it should make a difference for most users.

This uses Hamming windows:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'Hamming')
```

Hanning windows:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'Hanning')
```

Bartlett windows:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'Bartlett')
```

And here square (or 'rectangular') windows, which is equivalent to an un-edited window. These spectrograms clearly look worse:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'square')
```

Finally, this uses Blackman windows:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'Blackman')
```

::: {.callout-tip collapse="true"}
## Why should I care about window shape?

Typically you don't need to! The [Praat documentation](https://www.fon.hum.uva.nl/praat/manual/Advanced_spectrogram_settings___.html){target='_blank'} says that window shapes other than Gaussian are included for "pedagogical purposes" only. When plotting spectrograms of speech sounds, it typically won't matter much if you use Gaussian, Hanning, or Hamming windows, as they have very similar shapes. These windows look like this:

```{r}
#| fig-align: center

Gaussian_window <- phonTools::windowfunc(1000, type = 'gaussian')
Hann_window <- phonTools::windowfunc(1000, type = 'hann')
Hamming_window <- phonTools::windowfunc(1000, type = 'hamming')

plot(Gaussian_window, type = 'l', xaxt = 'n', yaxt = 'n', xlab = '', ylab = '',
     ylim = c(-0, 1), lwd = 2)
lines(Hann_window, col = 'blue', lwd = 2)
lines(Hamming_window, col = 'red', lwd = 2)
text(0, 1, labels = 'Gaussian', adj = 0)
text(0, 0.92, labels = 'Hanning', adj = 0, col = 'blue')
text(0, 0.84, labels = 'Hamming', adj = 0, col = 'red')
```
These are all variations on the same idea, smooth functions that allow most of the information at the center of a window to pass through and less information at the edges. Bartlett windows can also be considered a variation on the same general idea, only they are pyramid-shaped rather than smooth, while square windows simply allow all information to pass through. Blackman windows are also smooth -- these are not available in Praat, but are available in the `phonTools` library that are used to generate spectrograms in `praatpicture`.

```{r}
#| fig-align: center

Blackman_window <- phonTools::windowfunc(1000, type = 'blackman')
Bartlett_window <- phonTools::windowfunc(1000, type = 'bartlett')
square_window <- c(0, phonTools::windowfunc(998, type = 'rectangular'), 0)

plot(Blackman_window, type = 'l', xaxt = 'n', yaxt = 'n', xlab = '', ylab = '',
     ylim = c(-0, 1), lwd = 2)
lines(Bartlett_window, col = 'blue', lwd = 2)
lines(square_window, col = 'red', lwd = 2)
text(20, 0.95, labels = 'Blackman', adj = 0)
text(20, 0.87, labels = 'Bartlett', adj = 0, col = 'blue')
text(20, 0.79, labels = 'square', adj = 0, col = 'red')
```

The argument for relying on Gaussian window shapes in Praat is that they are supposedly the only window shape that doesn't produce so-called *sidelobes* when plotting a simple sine wave. A spectrogram showing a 200 Hz sine wave should show only a horizontal line near 200 Hz, but you may also see vertical stripes in the spectrogram that are an artefact of the window shape, and do not indicate anything present in the signal. For example, here is a spectrogram of a 1-second 200 Hz sine wave. 

```{r}
praatpicture('ex/simple.wav', frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'Hamming')
```

Unfortunately, when plotting spectrograms using the Gaussian window shapes available in the `phonTools` package, sidelobes are also visible in the spectrogram, although they are less prominent:

```{r}
praatpicture('ex/simple.wav', frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'Gaussian')
```

This does not happen in Praat, and must mean that the `phonTools` Gaussian window shapes are not the same as those used by Praat. This is a mystery I have yet to solve, and this is why `praatpicture` also supports Blackman windows, because they do not have these artefacts:

```{r}
praatpicture('ex/simple.wav', frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_windowShape = 'Blackman')
```

:::

### Time step

By default, spectrograms are composed of 1,000 spectral slices. This is controlled with the `spec_timeStep` argument. Lowering this will speed up generating the spectrogram, but the spectrogram will likely look pixellated. Here is one composed from "only" 200 spectral slices:

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_timeStep = 200)
```

500 spectral slices will usually be enough to render a spectrogram that doesn't look pixellated, and will also speed up generating the spectrogram a fair bit.

```{r}
#| fig-align: center

praatpicture('ex/tg.wav',
             frames = c('sound', 'spectrogram'),
             proportion = c(30,70),
             wave_color = 'grey',
             spec_timeStep = 500)
```

If you're printing a really large plot, that might not be enough though. 1,000 is enough for most purposes, but again, if you're printing something very large, you may want to set a higher time step.