# The waveform {#sec-wave}

```{r}
#| echo: false

library(praatpicture)
```


## Managing channels

`praatpicture()` has the option of plotting multiple channels. By default, all channels are plotted. If you have a stereo sound file, it will look like this:

```{r}
#| fig-align: center

praatpicture('ex/index_stereo.wav', 
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30))
```

You may not want this; if you prefer plotting just a single channel, you can controls this with the `wave_channels` argument. For plotting just the second channel:

```{r}
#| fig-align: center

praatpicture('ex/index_stereo.wav', 
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_channels = 2)
```

As indicated above, it is also possible plot audio with more than two channels, like in this case, where one channel captures audio, and other channels capture various articulatory signals:

```{r}
#| fig-align: center

praatpicture('ex/multichannel.wav', 
             start = 0.6, end = 2.4,
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30))
```

The argument `wave_channels` can be used to both extract and reorder channels. Here we omit the second channel, and swap the order of the third and fourth channels:

```{r}
#| fig-align: center

praatpicture('ex/multichannel.wav', 
             start = 0.6, end = 2.4,
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_channels = c(1,4,3))
```

## Labelling channels

The argument `wave_channelNames` can be used to add labels next to the y-axis with channel names. It is set to `FALSE` by default. If we plot stereo data with `wave_channelNames` set to `TRUE`, it'll by default print the labels "left" and "right" next to the channels: 

```{r}
#| fig-align: center

praatpicture('ex/index_stereo.wav', 
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_channelNames = TRUE)
```

Instead of passing a logical argument like `TRUE`, we can also pass a vector of strings and thus control the channel labels: 

```{r}
#| fig-align: center

praatpicture('ex/index_stereo.wav', 
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_channelNames = c('above', 'below'))
```

This is especially useful in cases like our multichannel data above, where we can then label the sources of the signals:

```{r}
#| fig-align: center

praatpicture('ex/multichannel.wav', 
             start = 0.6, end = 2.4,
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_channelNames = c('audio', 'airflow', 'EGG', 'Pio'))
```

You can of course also label just a single channel:

```{r}
#| fig-align: center

praatpicture('ex/index.wav', 
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_channelNames = 'wave')
```

## Coloring options

The argument `wave_color` allows users to control the color of the waveform. By default it is set to `black`; when plotting waveforms along with spectrogram, it is often the spectrogram in focus, so it can make a lot of sense to give the waveform a lighter grey color like so:

```{r}
#| fig-align: center

praatpicture('ex/index.wav', 
             wave_color='grey')
```

When multichannel data are plotted, it is also possible to assign different colors to different channels, by passing a vector instead of a single string to `wave_color`. 

```{r}
#| fig-align: center

praatpicture('ex/multichannel.wav', 
             start = 0.6, end = 2.4,
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_channelNames = c('audio', 'airflow', 'Pio', 'EGG'),
             wave_color=c('red', 'black', 'orange', 'blue'))
```

## Line width

The width of the waveform line can be controlled with the `wave_lineWidth` argument. If you're plotting a very complex wave, it can be helpful to pass a number smaller than `1`, which is the default:

```{r}
#| fig-align: center

praatpicture('ex/index.wav', 
             frames = c('sound', 'TextGrid'),
             proportion = c(70, 30),
             wave_lineWidth = 0.5)
```

You typically won't want to pass a number larger than `1`, but it could make sense if you're plotting only a very short sound snippet.